---
title: "Getting started with tinytrail"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tinytrail}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r load}
library(tinytrail)
```

`tinytrail` is a package that — once initialized — leaves a 'tiny trail' of
human- and AI-readable log-entries in a `yaml` that makes it effortless to keep
track of small to medium-sized projects. The package is lightweight (hence
'tiny') and maintains a YAML trail file recording which scripts produced which
output files.

## The two core functions

Place two calls in every script and the trail stays current automatically.

| Function | Where to use it |
|---|---|
| `tinytrail()` | Once, near the top of every script |
| `tinytrail_write()` | Inline around any file path you save to |

Call `tinytrail()` once at the top of each script with a short description of
what the script does:

```{r core, eval=FALSE}
library(tinytrail)

# --- top of 01_clean.R ---
tinytrail(
  description = "Clean and reshape survey data",
  data_source = "Current Population Survey (BLS)"
)
```

Then wrap the file path argument of any save call with `tinytrail_write()`.
It logs the path and returns it unchanged, so it slots directly into the
`file =` argument of any write function:

```{r write, eval=FALSE}
write.csv(
  survey,
  file = tinytrail_write("data/clean/survey_clean.csv")
)
```

It works equally well with `here::here()`:

```{r here, eval=FALSE}
write.csv(
  survey,
  file = tinytrail_write(here::here("data/clean/survey_clean.csv"))
)
```

That is all you need to add. The trail file (`_tinytrail.yaml`) is created in
the project root on the first run and updated on every subsequent run:

```yaml
$version: 0.1.0
$learn_more: https://github.com/tomasrei/tinytrail
scripts:
  01_clean.R:
    description: Clean and reshape survey data
    data_source: Current Population Survey (BLS)
    first_run: '2026-01-15 09:12'
    latest_run: '2026-01-20 11:05'
    script_runtime: 0.3 min
    n_outputs: 1
    outputs:
    - data/clean/survey_clean.csv
  02_model.R:
    description: Fit logistic regression
    data_source: Cleaned survey data (01_clean.R)
    first_run: '2026-01-16 14:22'
    latest_run: '2026-01-20 11:08'
    script_runtime: 1.4 min
    n_outputs: 2
    outputs:
    - results/model_fit.rds
    - results/table_coefficients.csv
```

## Adding a data dictionary

`tinytrail_dict()` is an optional third function. Place it at the end of a
read or clean pipeline to capture column names and sample values for each input
data frame:

```{r dict, eval=FALSE}
survey <- read.csv("data/raw/survey.csv") |>
  tinytrail_dict()
```

The entry is stored under the calling script in the trail file:

```yaml
data_dictionary:
  01_clean.R:
    survey:
      columns:
        id: [1, 2, 3, 4, 5]
        age: [34, 52, 28, 41, 37]
        response: ['yes', 'no', 'yes', 'yes', 'no']
```

To omit sample values and record only column names, use `sample_values = FALSE`:

```{r dict-no-samples, eval=FALSE}
survey <- read.csv("data/raw/survey.csv") |>
  tinytrail_dict(sample_values = FALSE)
```

## Options

Pin an important script to the top of the trail — useful for a `main.R` that
sources other scripts:

```{r pin, eval=FALSE}
tinytrail(
  description = "Master data preparation",
  data_source = "Admin registry (Statistics Sweden)",
  pin_to_top  = TRUE
)
```
