---
title: "Getting started with tinytrail"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tinytrail}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`tinytrail` is a package that — once initialised — leaves a 'tiny trail' of
human- and AI-readable text that makes it effortless to keep track of small to
medium-sized projects. The package is lightweight (hence 'tiny') and ...

`tinytrail` keeps a YAML registry of which scripts produced which output files
in your project. Place three lightweight calls in each script and the registry
stays current automatically.

## The three functions

| Function | Where to use it |
|---|---|
| `tinytrail()` | Top of every script |
| `tinytrail_write()` | Inline around any file path you save to |
| `tinytrail_dict()` | After reading/cleaning a data frame |

## A minimal workflow

```{r workflow}
library(tinytrail)

# --- top of 01_clean.R ---
tinytrail(
  data_source = "The Current Population Survey (BLS)",
  description = "Clean and reshape survey data"
)

# read and clean
survey <- read.csv("data/raw/survey.csv") |>
  tinytrail_dict()                           # records column names + samples

# save — wrap the file path with tinytrail_write()
write.csv(
  survey,
  file = tinytrail_write("data/clean/survey_clean.csv")
)
```

`tinytrail_write()` is a thin wrapper that logs the path and returns it unchanged,
so it slots directly into the `file =` argument of any write function. It works
equally well with `here::here()`:

```{r here}
write.csv(
  survey,
  file = tinytrail_write(here::here("data/clean/survey_clean.csv"))
)
```

That is all you need to add. The trail file (`_tinytrail.yaml`) is
created in the project root on the first run and updated on every subsequent
run.

## The trail file

After running a few scripts, `_tinytrail.yaml` looks like this:

```yaml
$version: 0.1.0
$learn_more: https://github.com/tomasrei/tinytrail
scripts:
  01_clean.R:
    data_source: data/raw/survey.csv
    description: Clean and reshape survey data
    first_run: '2026-01-15 09:12'
    latest_run: '2026-01-20 11:05'
    script_runtime: 0.3 min
    n_files: 1
    outputs:
    - data/clean/survey_clean.csv
  02_model.R:
    data_source: data/clean/survey_clean.csv
    description: Fit logistic regression
    first_run: '2026-01-16 14:22'
    latest_run: '2026-01-20 11:08'
    script_runtime: 1.4 min
    n_files: 2
    outputs:
    - results/model_fit.rds
    - results/table_coefficients.csv
data_dictionary:
  01_clean.R:
    survey:
      columns:
        id: [1, 2, 3, 4, 5]
        age: [34, 52, 28, 41, 37]
        response: ['yes', 'no', 'yes', 'yes', 'no']
```

## Options

Pin an important script to the top of the trail:

```{r pin}
tinytrail(
  data_source = "data/raw/master.csv",
  description = "Master data preparation",
  pin_to_top  = TRUE
)
```
